Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-09-24 intro.tex
commiting before deleting this file; it's my own cruft from the TFP work and isn't included anywhere
removed comments for the clean copy of intro, #6

First, it requires that programmers learn the subtleties of the textual syntax (e.g. the precedence of the various forms.) This can be particularly challenging for novices \cite{Altadmri:2015:MCI:2676723.2677258}. %For example, one study of novice programmers found syntax errors (e.g. unbalanced parentheses) to be the most common class of errors.
% Too much detail?

The fact that not every sequence of characters corresponds to a meaningful program also complicates matters for tool designers. In particular, program editors must contend with meaningless text on a regular basis. In a dataset analyzed by Yoon and Myers consisting of 1460 hours of edit logs, 44.2\% of edit states were syntactically malformed \cite{6883030}. Some additional percentage of edit states were well-formed but ill-typed (the dataset did not contain enough information to determine the exact percentage.)
Some of these meaningless edit states arose because the programmer was in the midst of a sequence of edits that left the edit state transiently meaningless, while others arose because the programmer made a logical mistake. Because the language definitions is silent about these edit states, it is difficult to design useful editor services, e.g. syntax highlighting~\cite{sarkar2015impact}, type-aware code completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324}, and refactoring support \cite{mens2004survey}. Editors must either disable these editor services when they encounter meaningless edit states or develop \emph{ad hoc} heuristics, which can be misleading. %, or because the programmer has made a mistake.
% The syntax definition is silent about these situations, so it is difficult for tools to help programmers determine and execute a corrective course of action

% Many editors have developed \emph{ad hoc} workarounds for this problem, e.g. they might attempt to use regular expressions to highlight malformed program text, insert closing delimiters automatically, use whitespace to guess where a delimiter is likely to appear, or continue past a type error by pretending that the type was as expected (if, indeed, an expected type can be determined.) These heuristic methods are often complex, and they can confuse or mislead the programmer. % This may also help explain why the error messages emitted by parsers and typecheckers are often quite baroque\todo{citations}.

% stuck editing a representation of the program instead of the structures
% themselves. The editor does not restrict what the programmer may do: you
% can delete characters that belong, insert ones that don't, forget things
% that were needed, and so. There's nothing stopping us from accidentally
% writing $$\lambda \mathit{x:num}.\mathit{(x,x)}$$ even though it's obvious
% that building a pair can't hope to form a natural number.

% The type structure of the language makes this sort of error obvious: it's
% not that you're adding characters that make your program incorrect, or even
% malformed; you're adding characters that can't possibly create a structure
% you want because of the type. Simply put, the primitive operations
% available in text editors do not always correspond to sensible
% transformations on the structure of the program.
These complications have motivated a long line of research into \emph{structure editors}, i.e. program editors where every edit state is a program structure. % Eliminating text eliminates the possibility of syntax errors.

@@

The Scratch language has a trivial static semantics, but researchers have also designed syntactic structure editors for  languages with a non-trivial static semantics. For example, \texttt{mbeddr} is an editor for a C-like language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an editor for a functional language similar to Haskell \cite{lamdu}. Each of these editors presents an innovative user interface, but the non-trivial type and binding structure of the underlying language complicates its design. The reason is that a syntactic structure editor does not guarantee that every edit state is statically meaningful -- only that it is syntactically well-formed. As in textual program editors, these syntactic structure editors must either disable key editor services when they encounter meaningless edit states or deploy \emph{ad hoc} heuristics.
%Indeed, it is not entirely clear what it should mean for terms with holes to be statically meaningful.
% Consequently, heuristics analagous to those that pervade textual program editors also pervade these syntactic structure editors.

% \begin{itemize}
% \item

% First, there is an all too familiar problem: these languages do not have a clear formal semantics and metatheory, i.e. ``the implementation is the specification''. This makes it difficult to reason methodically about types and binding, so, in turn, it is difficult to design reliable editor services.

% Even if a formal semantics and metatheory for the underlying language were forthcoming (e.g. following Standard ML \cite{Harper00atype-theoretic,mthm97-for-dart}), it would go only part of the way towards addressing the challenges faced by structure editor designers. First, a standard semantics assigns no formal meaning to \emph{incomplete terms}, i.e. terms with holes. Moreover, a syntactic structure editor  -- so incorporating holes into the static semantics would in any case be only a minor salve. Heuristics analagous to those that pervade textual program editors would still be needed to contend with the ill-typed edit states.

This paper develops a principled solution to this problem. We introduce Hazelnut,  a \emph{typed structure editor}  based on a bidirectionally typed lambda calculus extended to assign static meaning to expressions and types with {holes}, which we call \textbf{H-expressions} and \textbf{H-types}. Hazelnut's formal \emph{action semantics} maintains the invariant that every edit state is a statically meaningful (i.e. well-typed) H-expression with a single superimposed \emph{cursor}. We call H-expressions and H-types with a cursor \textbf{Z-expressions} and \textbf{Z-types} (so prefixed because our encoding follows Huet's \emph{zipper} pattern \cite{JFP::Huet1997}.) %ctions act relative to the H-expression or H-type under the cursor.% (which need not be a hole.)  %More specifically, actions are context-aware,  acting on an expression whose type is determined by its surroundings (e.g. a function argument), only actions consistent with that type are permitted.

Na\"ively, enforcing an injunction on ill-typed edit states would force programmers to construct programs in a rigid ``outside-in'' manner. For example, the programmer would often need to construct the outer function application form before identifying the intended function. To address this problem, Hazelnut leaves newly constructed expressions \emph{inside} a hole if the expression's type is inconsistent with the expected type. This meaningfully defers the type consistency check until the expression inside the hole is \emph{finished}. In other words, holes appear both at the leaves and at the internal nodes of an H-expression that remain under construction. %In short, Hazelnut is a \emph{bidirectionally typed structure editor calculus}. %Actions act at a programmer-indicate subtree, called the \emph{focus} (which is defined following Huet's zipper pattern.)

% By defining \emph{holes} as a language construct, for both expressions as types, Hazelnut enables type-aware actions that always leave the program in both a structurally and semantically well-defined state.
%In Hazelnut, expressions and types with \emph{holes} have a well-defined static semantics. Edit actions are type-aware and leave the program in both a structurally and semantically well-defined state.

%However, these syntactically correct states can be semantically meaningless (i.e. undefined), because the language definitions generally only give meaning to complete, well-typed terms. (e.g., when a branching element is introduced, the program does not become semantically correct until both branches are complete)
% This makes it difficult for humans and tools to reason about types and binding during the development process, even when using a structure editor.
% Similar to text editors, structure editors also can develop workarounds to try to help users, but these efforts can only extend so far, because of the underlying language definition.
%Some structure editors attempt stuff, but it is not clear what invariants are being maintained (e.g. Unison ; Scratch seems not to allow literals of the wrong type, but variables are not typed.)\todo{revise} %More sophisticated semantic reasoning principles thus remain .

The remainder of this paper is organized as follows:
\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
\item We begin in Sec.
    \ref{sec:example} with two examples of edit sequences to develop the reader's intuitions.
\item We then give a detailed overview of Hazelnut's semantics and metatheory, and our mechanization in the Agda proof assistant, in Sec.  \ref{sec:hazel}.
\item Hazelnut is a {foundational} calculus, i.e. a calculus that language and editor designers are expected to extend with higher level constructs. We extend Hazelnut with simple sum types in Sec.  \ref{sec:extending} to demonstrate how Hazelnut's rich metatheory guides such a development.
\item In Sec.  \ref{sec:impl}, we briefly describe how Hazelnut's action semantics lends itself to efficient implementation as a functional reactive program. Our reference implementation is written using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library.

% \item However, we consider the space of possible evaluation strategies for incomplete expressions in Section \ref{sec:dynamics}.
% so doing, we discover interesting connections with gradual typing and contextual modal type theory. The former provides an interpretation of type holes, and the latter provides a logical interpretation of expression holes. \todo{revise this one}

\item In Sec.  \ref{sec:rw}, we summarize related work. In particular, much of the technical machinery needed to handle type holes coincides with machinery developed in the study of gradual type systems. Similarly, expression holes can be interpreted as the closures of contextual modal type theory, which, by its correspondence with contextual modal logic, lays logical foundations beneath our work. %This suggests a principled design for an ``edit and resume'' feature.

>>>>

First, it requires that programmers learn the subtleties of the textual
syntax (e.g. the precedence of the various forms.) This can be particularly
challenging for novices \cite{Altadmri:2015:MCI:2676723.2677258}.

The fact that not every sequence of characters corresponds to a meaningful
program also complicates matters for tool designers. In particular, program
editors must contend with meaningless text on a regular basis. In a dataset
analyzed by Yoon and Myers consisting of 1460 hours of edit logs, 44.2\% of
edit states were syntactically malformed \cite{6883030}. Some additional
percentage of edit states were well-formed but ill-typed (the dataset did
not contain enough information to determine the exact percentage.)  Some of
these meaningless edit states arose because the programmer was in the midst
of a sequence of edits that left the edit state transiently meaningless,
while others arose because the programmer made a logical mistake. Because
the language definitions is silent about these edit states, it is difficult
to design useful editor services, e.g. syntax
highlighting~\cite{sarkar2015impact}, type-aware code
completion~\cite{Mooty:2010:CCC:1915084.1916348,Omar:2012:ACC:2337223.2337324},
and refactoring support \cite{mens2004survey}. Editors must either disable
these editor services when they encounter meaningless edit states or
develop \emph{ad hoc} heuristics, which can be misleading.

These complications have motivated a long line of research
into \emph{structure editors}, i.e. program editors where every edit state
is a program structure.

@@

The Scratch language has a trivial static semantics, but researchers have
also designed syntactic structure editors for languages with a non-trivial
static semantics. For example, \texttt{mbeddr} is an editor for a C-like
language \cite{voelter_mbeddr:_2012}, TouchDevelop is an editor for an
object-oriented language \cite{tillmann_touchdevelop:_2011} and Lamdu is an
editor for a functional language similar to Haskell \cite{lamdu}. Each of
these editors presents an innovative user interface, but the non-trivial
type and binding structure of the underlying language complicates its
design. The reason is that a syntactic structure editor does not guarantee
that every edit state is statically meaningful -- only that it is
syntactically well-formed. As in textual program editors, these syntactic
structure editors must either disable key editor services when they
encounter meaningless edit states or deploy \emph{ad hoc} heuristics.

This paper develops a principled solution to this problem. We introduce
Hazelnut, a \emph{typed structure editor} based on a bidirectionally typed
lambda calculus extended to assign static meaning to expressions and types
with {holes}, which we call \textbf{H-expressions}
and \textbf{H-types}. Hazelnut's formal \emph{action semantics} maintains
the invariant that every edit state is a statically meaningful
(i.e. well-typed) H-expression with a single superimposed \emph{cursor}. We
call H-expressions and H-types with a cursor \textbf{Z-expressions}
and \textbf{Z-types} (so prefixed because our encoding follows
Huet's \emph{zipper} pattern \cite{JFP::Huet1997}.)

Na\"ively, enforcing an injunction on ill-typed edit states would force
programmers to construct programs in a rigid ``outside-in'' manner. For
example, the programmer would often need to construct the outer function
application form before identifying the intended function. To address this
problem, Hazelnut leaves newly constructed expressions \emph{inside} a hole
if the expression's type is inconsistent with the expected type. This
meaningfully defers the type consistency check until the expression inside
the hole is \emph{finished}. In other words, holes appear both at the
leaves and at the internal nodes of an H-expression that remain under
construction.

The remainder of this paper is organized as follows:
\begin{itemize}[itemsep=0px,partopsep=2px,topsep=2px]
  \item We begin in Sec.  \ref{sec:example} with two examples of edit
    sequences to develop the reader's intuitions.

  \item We then give a detailed overview of Hazelnut's semantics and
  metatheory, and our mechanization in the Agda proof assistant, in
  Sec.  \ref{sec:hazel}.

  \item Hazelnut is a {foundational} calculus, i.e. a calculus that
  language and editor designers are expected to extend with higher level
  constructs. We extend Hazelnut with simple sum types in
  Sec.  \ref{sec:extending} to demonstrate how Hazelnut's rich metatheory
  guides such a development.

  \item In Sec.  \ref{sec:impl}, we briefly describe how Hazelnut's action
  semantics lends itself to efficient implementation as a functional
  reactive program. Our reference implementation is written
  using \lstinline{js_of_ocaml} and the OCaml \lstinline{React} library.

  \item In Sec.  \ref{sec:rw}, we summarize related work. In particular,
  much of the technical machinery needed to handle type holes coincides
  with machinery developed in the study of gradual type systems. Similarly,
  expression holes can be interpreted as the closures of contextual modal
  type theory, which, by its correspondence with contextual modal logic,
  lays logical foundations beneath our work.
